\texorpdfstring {Implementing Equational Constraints \\ in a Functional Language} {Implementing Equational Constraints in a Functional Language}

### 2012-11-14 papers/wlp2011/paper.tex
Comment added, bbl included

As mentioned in the introduction, the KiCS2 implementation
is based on the explicit representation of non-deterministic results
in a data structure.
This is achieved by extending each data type of the source program
by constructors to represent a choice between two values
and a failure, respectively.
For instance, the data type for Boolean values defined in a Curry program by
\begin{curry}
  data Bool = False | True
\end{curry}
is translated into the Haskell data type\footnote{Actually,
our compiler performs some renamings to avoid conflicts with
predefined Haskell entities and introduces type classes
to resolve overloaded symbols like \code{Choice} and \code{Fail}.}
\begin{haskell}
  data Bool = False | True | Choice ID Bool Bool | Fail
\end{haskell}
where \code{Fail} represents a failure and \code{(Choice i t1 t2)}
a non-deterministic value, i.e.,
a selection of two values \code{t1} and \code{t2}
that can be chosen by some search strategy.
The first argument \code{i} of type \code{ID} of a \code{Choice} constructor
is used to implement the call-time choice semantics
discussed in Sect.~\ref{sec:Curry}.
Since the evaluation of \code{xorSelf aBool} duplicates
the argument operation \code{aBool}, we have to ensure
that both duplicates, which later evaluate to a non-deterministic
choice between two values, yield either \code{True} or \code{False}.
This is obtained by assigning a unique identifier (of type \code{ID})
to each \code{Choice} constructor. The difficulty is to get a unique identifier
on demand, i.e., when some operation evaluates to a \code{Choice}.
We cannot thread an identifier supply, e.g. a counter,
through the search tree without fixing an evaluation order.
Since we want to compile into \emph{purely} functional programs
(in order to enable powerful program optimizations),
we can neither use unsafe features with side effects to generate
such identifiers. 
Hence, we follow the idea presented in \cite{AugustssonRittriSynek94} and
pass a (conceptually infinite) set of identifiers,
also called \emph{identifier supply},
to each operation so that a \code{Choice} can pick its unique identifier from
this set.
For this purpose, we assume a type \code{IDSupply},
representing an infinite set of identifiers,
with operations
\begin{haskell}
  initSupply  :: IO IDSupply
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
The operation \code{initSupply} creates such a set (at the beginning
of an execution),
the operation \code{thisID} takes some identifier from this set, and
\code{leftSupply} and \code{rightSupply} split this set
into two disjoint subsets without the identifier
obtained by \code{thisID}.
There are different implementations available
(see below for a simple one) and our system
is parametric over concrete implementations of \code{IDSupply}.

@@

A well-known disadvantage of equational constraints
is the fact that \ccode{=:=} is interpreted as strict equality.
Thus, if one uses equational constraints to express
requirements on arguments, the resulting operations might be too strict.
For instance, the equational constraint in the condition
defining \code{last} (see Sect.~\ref{ex:last})
requires that \code{ys++[e]} as well as \code{xs}
must be reducible to unifiable terms so that in consequence
the input list \code{xs} is completely evaluated.
Hence, if \code{failed} denotes an operation whose evaluation fails,
the evaluation of \code{last [failed,True]} has no result.
On the other hand, the evaluation of \code{last' [failed,True]} yields
the value \code{True}, i.e., the definition of \code{last'} is less strict
thanks to the use of functional patterns.

TODO fuer mh: Weitere Beispiele \cite{AntoyHanus05LOPSTR,Hanus11ICLP}

% As another example for the advantage of the reduced strictness
% implied by functional patterns,
% consider an operation that returns the first duplicate element in
% a list.
% Using equational constraints, we can define it as follows:
% \begin{curry}
%   fstDup xs | xs =:= ys++[e]++zs & elem e ys =:= True & nub ys =:= ys
%             = e    where ys, zs, e free
% \end{curry}
% The first equational constraint is used to split the input list \code{xs}
% into three sublists.
% The last equational constraint ensures that the first sublist
% \code{ys} does not contain duplicate elements
% (the library operation \code{nub} removes all duplicates from a list)
% and the second equational constraint ensures that the first element
% after \code{ys} occurs in \code{ys}.
% Although this implementation is concise, it cannot be applied
% to infinite lists due to the strict interpretation of \ccode{=:=}.
% This is not the case if we define this operation by a functional pattern:
% \begin{curry}
%   fstDup' (ys++[e]++zs) | elem e ys =:= True & nub ys =:= ys
%                         = e
% \end{curry}
% %
% Because of the reduced strictness, the logic variable \code{zs} (matching
% the tail list after the first duplicate) is never evaluated.

@@

\bibliographystyle{mhnumber}
\bibliography{mh}

>>>>

As mentioned in the introduction, the KiCS2 implementation
is based on the explicit representation of non-deterministic results
in a data structure.
This is achieved by extending each data type of the source program
by constructors to represent a choice between two values
and a failure, respectively.
For instance, the data type for Boolean values defined in a Curry program by
\begin{curry}
  data Bool = False | True
\end{curry}
is translated into the Haskell data type\footnote{Actually,
our compiler performs some renamings to avoid conflicts with
predefined Haskell entities and introduces type classes
to resolve overloaded symbols like \code{Choice} and \code{Fail}.}
\begin{haskell}
  data Bool = False | True | Choice ID Bool Bool | Fail
\end{haskell}
where \code{Fail} represents a failure and \code{(Choice i t1 t2)}
a non-deterministic value, i.e.,
a selection of two values \code{t1} and \code{t2}
that can be chosen by some search strategy.
The first argument \code{i} of type \code{ID} of a \code{Choice} constructor
is used to implement the call-time choice semantics
discussed in Sect.~\ref{sec:Curry}.
Since the evaluation of \code{xorSelf aBool} duplicates
the argument operation \code{aBool}, we have to ensure
that both duplicates, which later evaluate to a non-deterministic
choice between two values, yield either \code{True} or \code{False}.
This is obtained by assigning a unique identifier (of type \code{ID})
to each \code{Choice} constructor. The difficulty is to get a unique identifier
on demand, i.e., when some operation evaluates to a \code{Choice}.
We cannot thread an identifier supply, e.g., a counter,
through the search tree without fixing an evaluation order.
Since we want to compile into \emph{purely} functional programs
(in order to enable powerful program optimizations),
we can neither use unsafe features with side effects to generate
such identifiers. 
Hence, we follow the idea presented in \cite{AugustssonRittriSynek94} and
pass a (conceptually infinite) set of identifiers,
also called \emph{identifier supply},
to each operation so that a \code{Choice} can pick its unique identifier from
this set.
For this purpose, we assume a type \code{IDSupply},
representing an infinite set of identifiers,
with operations
\begin{haskell}
  initSupply  :: IO IDSupply
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
The operation \code{initSupply} creates such a set (at the beginning
of an execution),
the operation \code{thisID} takes some identifier from this set, and
\code{leftSupply} and \code{rightSupply} split this set
into two disjoint subsets without the identifier
obtained by \code{thisID}.
There are different implementations available
(see below for a simple one) and our system
is parametric over concrete implementations of \code{IDSupply}.

@@

A well-known disadvantage of equational constraints
is the fact that \ccode{=:=} is interpreted as strict equality.
Thus, if one uses equational constraints to express
requirements on arguments, the resulting operations might be too strict.
For instance, the equational constraint in the condition
defining \code{last} (see Sect.~\ref{ex:last})
requires that \code{ys++[e]} as well as \code{xs}
must be reducible to unifiable terms so that in consequence
the input list \code{xs} is completely evaluated.
Hence, if \code{failed} denotes an operation whose evaluation fails,
the evaluation of \code{last [failed,True]} has no result.
On the other hand, the evaluation of \code{last' [failed,True]} yields
the value \code{True}, i.e., the definition of \code{last'} is less strict
thanks to the use of functional patterns.
%
Beyond this improved operational behavior,
functional patterns can lead to more expressive programs
(e.g., matching and unification on infinite structures,
pattern matching at arbitrary depth in recursive data structures)
and more elegant program patterns
(see \cite{AntoyHanus05LOPSTR,AntoyHanus11WFLP,Hanus11ICLP} for examples).

@@

\begin{thebibliography}{10}

\bibitem{Ait-Kaci91}
H.~A{\"\i}t-Kaci.
\newblock {\em Warren's Abstract Machine}.
\newblock MIT Press, 1991.

\bibitem{AntoyHanus00FROCOS}
S.~Antoy and M.~Hanus.
\newblock Compiling Multi-Paradigm Declarative Programs into {Prolog}.
\newblock In {\em Proc. International Workshop on Frontiers of Combining
  Systems (FroCoS'2000)}, pp. 171--185. Springer LNCS 1794, 2000.

\bibitem{AntoyHanus02FLOPS}
S.~Antoy and M.~Hanus.
\newblock Functional Logic Design Patterns.
\newblock In {\em Proc.\ of the 6th International Symposium on Functional and
  Logic Programming (FLOPS 2002)}, pp. 67--87. Springer LNCS 2441, 2002.

\bibitem{AntoyHanus05LOPSTR}
S.~Antoy and M.~Hanus.
\newblock Declarative Programming with Function Patterns.
\newblock In {\em Proceedings of the International Symposium on Logic-based
  Program Synthesis and Transformation (LOPSTR'05)}, pp. 6--22. Springer LNCS
  3901, 2005.

\bibitem{AntoyHanus06ICLP}
S.~Antoy and M.~Hanus.
\newblock Overlapping Rules and Logic Variables in Functional Logic Programs.
\newblock In {\em Proceedings of the 22nd International Conference on Logic
  Programming (ICLP 2006)}, pp. 87--101. Springer LNCS 4079, 2006.

\bibitem{AntoyHanus10CACM}
S.~Antoy and M.~Hanus.
\newblock Functional Logic Programming.
\newblock {\em Communications of the ACM}, Vol.~53, No.~4, pp. 74--85, 2010.

\bibitem{AntoyHanus11WFLP}
S.~Antoy and M.~Hanus.
\newblock New Functional Logic Design Patterns.
\newblock In {\em Proc.\ of the 20th International Workshop on Functional and
  (Constraint) Logic Programming (WFLP 2011)}, pp. 19--34. Springer LNCS 6816,
  2011.

\bibitem{AntoyHanusLiuTolmach05}
S.~Antoy, M.~Hanus, J.~Liu, and A.~Tolmach.
\newblock A Virtual Machine for Functional Logic Computations.
\newblock In {\em Proc.\ of the 16th International Workshop on Implementation
  and Application of Functional Languages (IFL 2004)}, pp. 108--125. Springer
  LNCS 3474, 2005.

\bibitem{AugustssonRittriSynek94}
L.~Augustsson, M.~Rittri, and D.~Synek.
\newblock On generating unique names.
\newblock {\em Journal of Functional Programming}, Vol.~4, No.~1, pp. 117--123,
  1994.

\bibitem{BrasselFischer08IFL}
B.~Bra{\ss}el and S.~Fischer.
\newblock From Functional Logic Programs to Purely Functional Programs
  Preserving Laziness.
\newblock In {\em Proceedings of the 20th International Symposium on
  Implementation and Application of Functional Languages (IFL 2008)}, pp.
  25--42. Springer LNCS 5836, 2008.

\bibitem{BrasselHanusPeemoellerReck11}
B.~Bra{\ss}el, M.~Hanus, B.~Peem{\"o}ller, and F.~Reck.
\newblock {KiCS2}: A New Compiler from {Curry} to {Haskell}.
\newblock In {\em Proc. of the 20th International Workshop on Functional and
  (Constraint) Logic Programming (WFLP 2011)}, pp. 1--18. Springer LNCS 6816,
  2011.

\bibitem{BrasselHuch07}
B.~Bra{\ss}el and F.~Huch.
\newblock On a Tighter Integration of Functional and Logic Programming.
\newblock In {\em Proc. APLAS 2007}, pp. 122--138. Springer LNCS 4807, 2007.

\bibitem{BrasselHuch09}
B.~Bra{\ss}el and F.~Huch.
\newblock The {Kiel} {Curry} {System} {KiCS}.
\newblock In {\em Applications of Declarative Programming and Knowledge
  Management}, pp. 195--205. Springer LNAI 5437, 2009.

\bibitem{GiovannettiLeviMoisoPalamidessi91}
E.~Giovannetti, G.~Levi, C.~Moiso, and C.~Palamidessi.
\newblock Kernel LEAF: A Logic plus Functional Language.
\newblock {\em Journal of Computer and System Sciences}, Vol.~42, No.~2, pp.
  139--185, 1991.

\bibitem{GonzalezEtAl99}
J.C. Gonz{\'a}lez-Moreno, M.T. Hortal{\'a}-Gonz{\'a}lez, F.J.
  L{\'o}pez-Fraguas, and M.~Rodr{\'\i}guez-Artalejo.
\newblock An approach to declarative programming based on a rewriting logic.
\newblock {\em Journal of Logic Programming}, Vol.~40, pp. 47--87, 1999.

\bibitem{Hanus07ICLP}
M.~Hanus.
\newblock Multi-paradigm Declarative Languages.
\newblock In {\em Proceedings of the International Conference on Logic
  Programming (ICLP 2007)}, pp. 45--75. Springer LNCS 4670, 2007.

\bibitem{Hanus11ICLP}
M.~Hanus.
\newblock Declarative Processing of Semistructured Web Data.
\newblock In {\em Technical Communications of the 27th International Conference
  on Logic Programming}, volume~11, pp. 198--208. Leibniz International
  Proceedings in Informatics (LIPIcs), 2011.

\bibitem{Hanus10PAKCS}
M.~Hanus, S.~Antoy, B.~Bra{\ss}el, M.~Engelke, K.~H{\"o}ppner, J.~Koj,
  P.~Niederau, R.~Sadre, and F.~Steiner.
\newblock {PAKCS}: The {P}ortland {A}achen {K}iel {C}urry {S}ystem.
\newblock Available at \url{http://www.informatik.uni-kiel.de/~pakcs/}, 2010.

\bibitem{HanusPeemoellerReck12ATPS}
M.~Hanus, B.~Peem{\"o}ller, and F.~Reck.
\newblock Search Strategies for Functional Logic Programming.
\newblock In {\em Proc. of the 5th Working Conference on Programming Languages
  (ATPS'12)}, pp. 61--74. Springer LNI 199, 2012.

\bibitem{HanusSadre99JFLP}
M.~Hanus and R.~Sadre.
\newblock An Abstract Machine for {Curry} and its Concurrent Implementation in
  {Java}.
\newblock {\em Journal of Functional and Logic Programming}, Vol.~1999, No.~6,
  1999.

\bibitem{Hanus12Curry}
M.~Hanus~(ed.).
\newblock Curry: An Integrated Functional Logic Language (Vers.\ 0.8.3).
\newblock Available at \url{http://www.curry-language.org}, 2012.

\bibitem{Hussmann92}
H.~Hussmann.
\newblock Nondeterministic Algebraic Specifications and Nonconfluent Term
  Rewriting.
\newblock {\em Journal of Logic Programming}, Vol.~12, pp. 237--255, 1992.

\bibitem{Lopez-FraguasSanchez-Hernandez99}
F.~L\'opez-Fraguas and J.~S\'anchez-Hern\'andez.
\newblock {TOY}: A Multiparadigm Declarative System.
\newblock In {\em Proc. of RTA'99}, pp. 244--247. Springer LNCS 1631, 1999.

\bibitem{Lux99FLOPS}
W.~Lux.
\newblock Implementing Encapsulated Search for a Lazy Functional Logic
  Language.
\newblock In {\em Proc. 4th Fuji International Symposium on Functional and
  Logic Programming (FLOPS'99)}, pp. 100--113. Springer LNCS 1722, 1999.

\bibitem{Naish85SLP}
L.~Naish.
\newblock All Solutions Predicates in Prolog.
\newblock In {\em Proc. IEEE Internat. Symposium on Logic Programming}, pp.
  73--77, Boston, 1985.

\bibitem{ODonnell85}
M.J. O'Donnell.
\newblock {\em Equational Logic as a Programming Language}.
\newblock MIT Press, 1985.

\bibitem{PeytonJones03Haskell}
S.~Peyton~Jones, editor.
\newblock {\em Haskell 98 Language and Libraries---The Revised Report}.
\newblock Cambridge University Press, 2003.

\bibitem{Reddy85}
U.S. Reddy.
\newblock Narrowing as the Operational Semantics of Functional Languages.
\newblock In {\em Proc. IEEE Internat. Symposium on Logic Programming}, pp.
  138--151, Boston, 1985.

\bibitem{Robinson65}
J.A. Robinson.
\newblock A Machine-Oriented Logic Based on the Resolution Principle.
\newblock {\em Journal of the ACM}, Vol.~12, No.~1, pp. 23--41, 1965.

\bibitem{SchrijversStuckeyWadler09}
T.~Schrijvers, P.~Stuckey, and P.~Wadler.
\newblock Monadic Constraint Programming.
\newblock {\em Journal of Functional Programming}, Vol.~19, No.~6, pp.
  663--697, 2009.

\bibitem{Slagle74}
J.R. Slagle.
\newblock Automated Theorem-Proving for Theories with Simplifiers,
  Commutativity, and Associativity.
\newblock {\em Journal of the ACM}, Vol.~21, No.~4, pp. 622--642, 1974.

\end{thebibliography}
